---
title: "FISH 627: Homework 3"
author: "Casey Clark"
date: "Due by end of Thursday, October 22, 2015"
output: html_document
---

<span style="color:red">**Total: 10 points**  
Out of fairness to others (and to me) I cannot accept late homework. I will subtract 0.5 points per day after the due date and will not accept homework that is more than a week late! If you have valid reasons such as field work, conferences, etc. that prevent you from turning in a homework on time, talk to me beforehand!</span>  

#Fitting a LOWESS model and plotting results:  
For this problem we will estimate average ozone values at different levels of wind and temperature (as well as different levels of solar radiation) and plot the resulting estimates as a smooth surface. 

**a.** 
<span style="color:red">(1 pt.)</span>  Attach the *lattice* library and explore the ozone data in data frame "environmental". For simplicity, you may also want to attach the environmental data frame, but make sure you enderstand what happens when you do! 
```{r load, message=FALSE}
library(lattice)
#?environmental
attach(environmental) #attach can induce some issues so we will detach asap
enviro <- environmental
detach(environmental)
```

Plot a histogram of ozone using *'hist'*. Compare a histogram of these "raw" ozone values to one that uses 3^rd^ root transformed values (ozone^(1/3)). <span style="color:blue"> Describe the difference and explain
why we would want to transform the ozone variable if used as dependent variable in a typical regression (this requires some familiarity with standard regression assumptions).</span> 

```{r a}
library(lattice)
?environmental
attach(environmental)

hist(ozone) #These data are not normal.  
            #Large positive skew (lots of values near minimum, long tail out to right)

hist(ozone^(1/3)) #The 3rd root transformation brings the data much closer to normality
                  #The histogram approximates a bell curve and the mean has moved away
                  #from the minimum value. This helps meet the normality assumption
                  #required for many parametric statistics
```



**b.**
<span style="color:red">(0.5 pt.)</span> Fit a LOWESS model of (transformed) ozone levels on wind and temperature using the 'loess' function, similar to the model for GAK 1 temperatures that we fit in Module 8:
```{r b, eval=FALSE}
ozo.trans <- ozone^(1/3) #make an object of the transformed ozone data
ozo1 <- loess(ozo.trans ~ wind + temperature, span = 1, degree = 2) 
                          #Fit a LOWESS model to the transformed ozone data on
                          #wind and temperature using the suggested settings
```

Use the suggested setting and insert the appropriate *formula*.

**c.**
<span style="color:red">(1 pt.)</span> The LOWESS fit does not have regression coefficients like a linear regression model to compute predicted values, but you can extract the fitted values and residuals from the fitted model object (```ozo1```) using functions ```fitted``` and ```resid```. Plot the observed values (on the 
transformed scale!!) against the fitted values. Add a 1:1 line to the plot (intercept 0, slope 1) that 
would denote a perfect fit (i.e. a 1:1 correspondence):  
```abline(0,1)```
```{r c}
fitted(ozo1) #extract and look at fitted values
resid(ozo1)  #extract and look at residuals

plot(ozo.trans ~ fitted(ozo1)) #plot the transformed ozone values against the fitted values
abline(0,1) #add a 1:1 line. The relationship is pretty close to 1:1.
```

**d.**
<span style="color:red">(1 pt.)</span>  The LOWESS model is similar to any linear regression in that it finds fitted values by minimizing the sum of squared differences between the observations and the fitted regression line or (in this case) the fitted 2-dimensional surface (as in "least-squares" regression, although LOWESS differs in that it uses a modified least-squares criterion). The model fit can be 
evaluated by examining the residual variability, which is described by the spread of the observed 
values around the fitted values and is typically quantified by the 'residual standard error'.  

You can view the residual standard error in the summary output using:  
```summary(ozo1)```
```{r d}
summary(ozo1)   #Residual standard error = 0.501

p <- summary(ozo1)$enp  #extract the expected number of parameters
n <- summary(ozo1)$n    #extract the number of observations

resid.se <- sqrt(((sum(resid(ozo1)^2)))/(n - p))  #directly calculate RSE
resid.se     # = 0.497  pretty close to the RSE given in the summary
```

For comparison, compute the standard error directly from the residuals. The residual standard error can be estimated as:  

$$	\hat{\sigma}_{\varepsilon}^2=\sqrt{\sum\limits_{i=1}^n\sigma_{i}^2\Big/(n-p)} $$  

All the information you need is either in the `summary` output (n is the number of observations
and p is the equivalent number of parameters) or can be extracted from the fitted object (i.e. 
```resid(ozo1)``` returns the vector of residuals or ε). You can use vectorized arithmetic, that is you 
can square the entire residual vector ε and then sum over the elements εi using`'sum`. There is no 
need for looping (!!), and the above formula can be written as a single short expression using 
`sqrt`, `sum`, and basic arithmetic operators. You can take the values for *n* and *p* from 
'summary(ozo1)' and simply type them into your expression to compute the standard error or 
you can extract n and p from the summary object and then use them in your expression:  
```p <- ozo1$enp```   
```n <- ozo1$n```  

Use ```str(ozo1)``` to examine all available elements.  
(Note that the computed value for the standard error of the residuals differs somewhat from the 
value returned in the summary output due to slight differences in the way they are calculated).

**e.**
<span style="color:red">(0.5 pt.)</span> Remember that any regression model "decomposes" the data into predicted values 
(the fitted values you extracted earliler) plus the residual variability (the residuals you extracted 
earlier), that is: $y = \hat{y} + \hat{\varepsilon}$ where $y$ is the data vector, $\hat{y}$  is a vector of fitted values, and $\hat{\varepsilon}$ is a vector of residuals.   

Verify that the sum of the fitted values and the residuals is equal to the observed ozone values
(e.g. by plotting (fitted values + residual values) against (observed values) or by computing the 
differences, which should all be 0).
```{r e}
res.p.fit <- resid(ozo1) + fitted(ozo1) #Sum the vectors of the residuals and fitted values
plot(ozo.trans ~ res.p.fit)  #Plot the summed values against the observed values 
abline(0,1) #Perfect 1:1 line (i.e., the values are identical)

ozo.trans - res.p.fit  #Can also double check by computing the differences (which = 0)
```

**f.**
<span style="color:red">(1.5 pt.)</span> Examine the fit of the model by doing some basic residual diagnostics / plots. Split the graphics window into 2 rows and 2 columns to examine all plots at once: 
```par(mfrow=c(2,2))```  
Create the following set of plots, which should be part of model checking whenever you fit a 
statistical model: 

1. Plot residuals (on y-axis) against fitted values, add a horizontal line at 0 (```abline(h=0)```)
2. Plot residuals against wind and add line at 0
3. Plot residuals against temperature and add line at 0
4. Create a ```qqplot``` that compares the quantiles of the residual distribution against the 
corresponding quantiles of a normal distribution. If residuals are normally distributed, they 
should roughly fall on a straight line:   
   ```qqnorm(resid(ozo1))```  
   ```qqline(resid(ozo1))```
```{r f}
par(mfrow = c(2,2))               #set up the graphics window for 4 panels

#1.
plot(resid(ozo1) ~ fitted(ozo1))  #plot residuals against fitted values
abline(h=0)                       #add horizontal line at zero

#2.
plot(resid(ozo1) ~ wind)          #plot residuals against wind
abline(h=0)                       #add horizontal line at zero

#3
plot(resid(ozo1) ~ temperature)   #plot residuals against temperature
abline(h=0)                       #add horizontal line at zero

#4
qqnorm(resid(ozo1))               #qqplot to check for normality.
qqline(resid(ozo1))

#The assumptions we can examine using these plots are as follows:
#   Independence of variables: Appears to be met.  There are not obvious patterns when the
#                              residuals are plotted against fitted values/variables                            
#   Equality of variances: Mostly looks ok.  The variability looks a little lower on the right
#                          side of the residuals ~ wind plot.  Maybe also in the 
#                          residuals ~ fitted plot?
#   Normality: This looks pretty good.  The values dip a little off the line at the extremes,
#              but overall not concerning

#So, in summary, it looks like the assumptions are probably met (or close enough)
```

<span style="color:blue"> Briefly discuss whether the regression assumptions are met: There should be no obvious patterns in the residuals (i.e. they should be randomly distributed around the zero line), no 
heteroscedasticity (i.e. variability should be roughly constant across different fitted values and 
different values of the covariates), and they should be approximately normally distributed.</span> 

**g.**
<span style="color:red">(1.5 pt.)</span> 
Next, we set up a grid to compute predicted values on a regularly spaced grid for the purpose of plotting the predicted model surface (which provides an estimate of how ozone levels vary with wind speed and temperature):

* Use ```seq``` to create a regularly spaced sequence of numbers from the minimum wind
measurement to the maximum and from the minimum temperature to the maximum. Use 
about 50 equally spaced values (i.e. ```length.out=50```)  
```wnd <- seq(..., ..., ...) ```  
```temp <- seq(..., ..., ...)```  
* Create a matrix that contains a combination of each value in 'wnd' and each value in
```temp```:
```grd <- expand.grid(list(wind=wnd, temperature = temp))```  
Examine ```grd``` to understand its structure!!

* Compute predicted values: Use the resulting matrix (```grd```) to compute predicted values
and then to plot them. Note that the names of the variables in ```grd``` MUST match the 
names of the variables that we used in fitting the model! You can then use the "new" 
wind and temperature values to compute the predicted values at each combination of 
wind and temperature in ```grd```:  
```ozo1.fit <- predict(ozo1, newdata = grd)```  
Examine the structure of the resulting matrix and compare its dimensions to the length of  
```wnd``` and ```temp```.

* Following the example in Module 8, create an image plot of the fitted ozone values (on
transformed scale) against wind and temperature:  
```image(x=,yu=, z=, ...)``` Add appropriate axis labels & title.  
 Use ```col=terrain.colors(50)```
 
 * Use the ```contour``` function to add contours to the image plot. ```contour``` takes the same
x,y,z values as ```image```, but be sure to set ```add=T``` to add contours to the existing plot 
(otherwise you get a new plot with contours only).

*  Finally, use ```points``` to show where the actual measurements were taken as we did in Module 9. 

<span style="color:blue"> Comment on the ability of the model to estimate ozone levels at a wind speed of 20mph when temperatures are in the 80s or 90s!</span> 
```{r g}
par(mfrow=c(1,1))  #Return graphics frame to one panel

wnd <- seq(from = min(wind), to = max(wind), length.out = 50)
                   #Create a vector of values spanning the range of the wind values
temp <- seq(from = min(temperature), to = max(temperature), length.out = 50)
                   #Create a vector of values spanning the range of the temperature values

grd <- expand.grid(wind = wnd, temperature = temp)  #Combine the values into a matrix (2D surface)

ozo1.fit <- predict(ozo1, newdata = grd) #Create predicted values for all points across the surface

image(wnd, temp, ozo1.fit, col = terrain.colors(50)) #create an image with wind on the
        #x axis, temperature on the y axis, and predicted ozone values as the z values
contour(wnd, temp, ozo1.fit, add=T)  #add contour lines

points(wind, temperature, pch = 16, cex=1)  #plot the points where measurements were taken

#The actual data tend towards the middle of the plot with very few data points for
#weak winds/low temperatures and strong winds/high temperatures. Thus, the bottom left
#and top right corners of the surface may be suspect.  This appears to be especially true
#for strong winds (~20 mph) at high temps (80-90 degrees), where contours show a steep increase
```

**h.**
<span style="color:red">(0.5 pt.)</span> 
Compare your image plot to an alternative solution that uses the lattice function
'contourplot', which automatically adds a legend. It works somewhat differently and we need 
to add the fitted values to the 'grd' matrix in order to use it:
```{r h, eval=FALSE}
grd[, "fit"] <- c(predict(ozo1, grd)) #Add the fitted values to the matrix
contourplot(fit ~ wind + temperature, data = grd, 
            cuts = 10, region = TRUE, xlab = "Wind Speed (mph)",
            ylab = "Temperature (F)", main = "Cube Root Ozone (cube root ppb)",
            col.regions = terrain.colors(50))

#I like the first plot better.  The transitions appear smoother and less blocky. This
#plot still has troubles in the top right corner.  I like that it has a scale.
```

**i.**
<span style="color:red">(1.5 pt.)</span> 
Extend the model with two independent variables ```ozo1``` to include an additional
covariate - solar radiation - which may also have an effect on ozone levels. The model fits a 
"hyper-surface" to the data in 3 dimensions. You can think of it as a multiple regression of ozone 
on three independent variables with all possible interactions. (where the "interactions" are a bit 
more flexible in a LOWESS model than in a linear model). The model essentially allows for a 
different ozone response to temperature and wind conditions at different levels of solar radiation:

```{r i}
ozo2 <- loess(ozo.trans ~ wind + temperature + radiation, span = 1, degree = 2)
summary(ozo2)  #The number of observations has stayed the same and the equivalent number
               #of parameters has increased (as would be expected)

anova(ozo1, ozo2)

#The F-test indicates a reduction in the RSS of 5.959, with an increase of 4.49
#"equivalent parameters"

#The probability we would get an F value as large as the observed one under the null
#hypothesis is very small < 0.07%.  Thus, the more complex model appears to be better.
```

Use an approximate F-test to compare the model fits:  
```anova(ozo1, ozo2)```

This is equivalent to comparing two nested linear models. From regression you should remember 
that nested models can be compared by evaluating the reduction in residual sum of squares 
(RSS) that results from fitting a more complex model (e.g. by adding another covariate in the 
model). The 'anova' function is used to carry out this F-test (which is only approximate in this 
case because this is not a linear model).

<span style="color:blue">How much did the residual sum of squares decrease in the more complex model (ozo2) compared to the simpler model (ozo1)? How many extra "equivalent parameters" were 
needed to achieve that reduction?</span> 

The F-ratio computes the relative reduction in RSS between the two models, standardized by the 
difference in number of parameters. The better the more complex model fits (i.e. the lower its 
RSS) the larger the F-value becomes and the more reason we have to reject the simpler model. It 
can be shown that F has an "F distribution" under the assumption that the simpler model is true, 
thus we can evaluate the probability that F is as large as or larger than the observed F using the 
theoretical F distribution (which can easily be computed and used to be tabulated in older Stats 

<span style="color:blue"> From the output above, what is the probability that we would get an F value as large as the observed one under the "null hypothesis" (which states that the simpler model is the "true" model)? What can you conclude about the models (i.e. which one appears to be the 
better model)?</span> 

**j.**
<span style="color:red">(1 pt.)</span> 
pt) Lastly, set up a 3-D grid with 4 different levels of radiation, compute predicted values,
and plot the results as before, but with a different fitted surface at each of 4 radiation levels:
```rad <- seq(min(radiation), max(radiation), length = 4)```  
```grd <- expand.grid(list(wind=wnd, temperature = temp, radiation=rad))```  
Examine ```grd``` to understand its structure!

Compute predicted values and add them to the matrix:  
```pred <- predict(ozo2, grd)```  
Examine ```pred``` (using ```str```, ```dim```, etc) to understand its structure! It returns an array of predicted values from our regression model with one predicted value at each combination of wind speed, 
temperature, and radiation level! 

The next line of code converts the array to a long vector and appends it to the ```grd``` matrix:   
```grd[, "fit"] <- pred```

Using ```c()``` converts the array to a matrix (simply type ```c(pred)``` at the prompt to look at the 
result). You would also use: ```as.vector(pred)```. If you didn't know (or trust me) that this works 
properly, you would have to carefully check that the predicted values in the resulting vector 
correspond to the rows in ```grd```. Of course, ```expand.grid()``` was designed to make sure that it 
works with ```predict()```.

Examine ```grd``` again to understand what you just did!  
Create a plot of the fitted values at different combinations of wind and temperature and at four 
different levels of radiation:

```{r j, eval=FALSE}
rad <- seq(min(radiation), max(radiation), length = 4)  #create a vector of 4 radiation values
grd <- expand.grid(list(wind = wnd, temperature = temp, radiation = rad))
    #Add radiation values to the grid

pred <- predict(ozo2, grd) #Predict ozone values based on ozo2 model
grd[,"fit"] <- c(pred) #Add them to the grid. I think the .pdf has a typo here (no "c()")

contourplot(fit ~ wind * temperature | radiation, data = grd,
            cuts = 10, region = TRUE, xlab = "Wind Speed (mph)",
            ylab = "Temperature (F)", main = "Cube Root Ozone (cube root ppb)",
            col.regions = terrain.colors(50))

#Estimated ozone levels are lowest at the lowest level of solar radiation (plot is dominated
#by green), then increase with increasing radiation (green shrinks, yellow and orange become
#more apparent) and is highest at low wind speeds and high temperatures (left side and top
#of plot)
```

In the resulting figure, the solar radiation level increases from the lower left panel to the upper 
right panel as indicated by the orange vertical bar in the strips at the top of each panel.   
<span style="color:blue">How do the estimated ozone levels differ at different levels of radiation?</span> 

**Code used to geneerate this document**  
https://github.com/ben-williams/ReproResearch/blob/master/hwk3.Rmd

